Joint estimation of phylogenies and ancestral states is not consistent

### 2018-01-02 joint_inf.tex
address comments
:woot: :woot: :woot:
Main text wordsmithing.

\begin{abstract}
Maximum likelihood is a widely used and well understood approach to estimate the phylogenetic topology and branch length given observed character data.
Any phylogenetic maximum likelihood solution must propose a means of handling unknown ancestral states.
Classically, these unknown intermediate states are averaged out, which leads to consistent estimation of the topology and continuous model parameters.
Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
While this method of joint maximum likelihood estimation can obtain estimates more efficiently, it has the potential to result in undesired properties of the estimates.
In this paper, we show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
We derive a region where generating data under a four-taxon tree results in a joint estimation procedure converging on the wrong topology.
If the topology is known a priori, there are similar cases where the branch lengths are consistently overestimated in the case of infinite-length sequences.
Although joint maximum likelihood may turn out to be a useful tool in practice, care should be taken when using this approach.
\end{abstract}

@@

Existing consistency proofs for maximum likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
In this case those additional parameters come in the form of the states of ancestral sequences.
Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$, their work motivates understanding when exactly this approximate technique breaks down, even when sequence data is generated without sampling error according to the exact model used for inference.

In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model and imagine data being generated on the four-taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known and fixed.

@@

To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the Farris topology.
We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology than for the true, generating Farris topology.
To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the Farris topology as a function of $x^*$ and $y^*$, and, similarly, we construct a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
When $C_0(x^*, y^*) < C_1(x^*, y^*)$, the likelihood in the Felsenstein case is larger than the likelihood in the Farris case, demonstrating inconsistency (Fig.~\ref{fig:inconsistency-farris}).
\begin{restatable}{theorem}{topoInconsist}
Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x, y, x, y, w\}$.
There exist $C_0(x^*, y^*), C_1(x^*, y^*)$ and a set of $0 < x^*, y^* < 1$ such that
$$
\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t) \le C_0(x^*, y^*),
$$
$$
C_1(x^*, y^*) \le \max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t)
$$
with $C_0(x^*, y^*) < C_1(x^*, y^*)$.
\end{restatable}
The proof of this theorem is by a detailed examination of inequalities.
The only intuition we can offer is that $\tau_2$ is favored in performing joint maximum likelihood since the likelihood for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its likelihood while $\tau_2$ has many more.
This enables more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.

@@

The general case is more challenging to analyze and so we will get somewhat weaker bounds.
In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
We show that there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
We fix bounds on these estimates and show that, in the general case, joint estimation either estimates $\hat{w}$ to be 1 or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
\begin{restatable}{theorem}{generalBranchInconsist}
\label{thm:general-bl}
Define $\gamma(x, y) = 4xy$.
For
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*),
$$
bounds
$$
\gamma_L := \gamma_{L}(x^*, y^*) \le \gamma(\hat{x}, \hat{y}),
$$
$$
\gamma_U := \gamma_{U}(x^*, y^*) \ge \gamma(\hat{x}, \hat{y}),
$$
and
$$
\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}),
$$
the maximum likelihood value $\hat{w}=1$ when
$$
%EM It's a little surprising that "naked" beta shows up here. While it seems like a correct statement, why can we have beta here to make the plots, while in another term we have to bound it with beta_L? Seems like we'd need a beta_U.
-\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
$$
\end{restatable}
This theorem can be used to show incorrect branch parameter estimates as follows.
If we don't want to allow any error in pendant branches, we can use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section (Fig.~\ref{fig:bl-inconsistency}).
For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
$$
x^*-x_{L} \le \hat{x} \le x^*+x_{U}
$$
and
$$
y^*-y_{L} \le \hat{y} \le y^*+y_{U}.
$$
Then useful $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.
As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
In this case, joint inference will either estimate $\hat{w}$ to have fidelity 1 or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.

@@

Previous work in phylogenetics has developed consistency counterexamples using the same four-taxa topologies used here \cite{Felsenstein1978-rr}.
In this previous work, when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more optimal when performing a particular estimation procedure.
This is the converse of what we have shown for joint inference, where the Felsenstein topology is more likely than the Farris topology.
Moreover, the inconsistency in \cite{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
This is not the case here; for our case, the inconsistency generally occurs when all branches are long, and has more to do with the choices of the form of the likelihood than from the interplay of long and short branches.
Difficulties in phylogenetic estimation when generating data on the ``Farris zone'' tree have been found by \cite{Siddall1998-hq}, though \cite{Swofford2001-hr} show that sequence length plays a major role in these issues.

>>>>

\begin{abstract}
Maximum likelihood is a widely used and well understood approach to estimate phylogenetic topology and branch length given observed character data.
Any phylogenetic maximum likelihood solution must propose a means of handling unknown ancestral states.
Classically, these unknown intermediate states are averaged out, which leads to consistent estimation of the topology and continuous model parameters.
Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
While this method of joint maximum likelihood estimation can obtain estimates more efficiently, its properties as an estimator are not yet clear.
In this paper, we show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
We derive a region where generating data under a four-taxon tree results in a joint estimation procedure that gives incorrect topological inferences in the limit of infinite-length sequences.
For branch length estimation on a correct topology, there are similar cases where the branch lengths are consistently overestimated in the case of infinite-length sequences.
Thus although joint maximum likelihood may turn out to be a useful tool in practice, care should be taken when using this approach.
\end{abstract}

@@

Statistical consistency means that an estimator gives the generating model with probability 1 in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
In this case those additional parameters come in the form of the states of ancestral sequences.
Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$, their work motivates understanding when exactly this approximate technique breaks down, even when sequence data is generated without sampling error according to the exact model used for inference.

In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model with data being generated on the four-taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known and fixed.

@@

To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the Farris topology (Fig.~\ref{fig:farris-fels-top}a).
We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology than for the true, generating Farris topology.
To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the Farris topology as a function of $x^*$ and $y^*$, and, similarly, we construct a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
When $C_0(x^*, y^*) < C_1(x^*, y^*)$, the likelihood in the Felsenstein case is larger than the likelihood in the Farris case, demonstrating inconsistency (Fig.~\ref{fig:inconsistency-farris}).
\begin{restatable}{theorem}{topoInconsist}
Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x, y, x, y, w\}$.
There exist $C_0(x^*, y^*), C_1(x^*, y^*)$ and a set of $0 < x^*, y^* < 1$ such that
$$
\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t) \le C_0(x^*, y^*),
$$
$$
C_1(x^*, y^*) \le \max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t)
$$
with $C_0(x^*, y^*) < C_1(x^*, y^*)$.
\end{restatable}
The proof of this theorem is by a detailed examination of inequalities.
The only intuition we can offer is that $\tau_2$ is favored in performing joint maximum likelihood since the likelihood for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its likelihood while $\tau_2$ has many more.
This enables more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.

@@

The general case is more challenging to analyze and so we will get somewhat weaker bounds.
In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
We show that there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
We fix bounds on these estimates and show that, in the general case, joint estimation either estimates $\hat{w}$ to be 1 or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
\begin{restatable}{theorem}{generalBranchInconsist}
\label{thm:general-bl}
Define $\gamma(x, y) = 4xy$.
For
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*),
$$
bounds
$$
\gamma_L := \gamma_{L}(x^*, y^*) \le \gamma(\hat{x}, \hat{y}),
$$
$$
\gamma_U := \gamma_{U}(x^*, y^*) \ge \gamma(\hat{x}, \hat{y}),
$$
and
$$
\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}),
$$
the maximum likelihood value $\hat{w}=1$ when
$$
-\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
$$
\end{restatable}
This theorem can be used to show incorrect branch parameter estimates as follows.
If we don't want to allow any error in pendant branches, we can use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section (Fig.~\ref{fig:bl-inconsistency}).
For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
$$
x^*-x_{L} \le \hat{x} \le x^*+x_{U}
$$
and
$$
y^*-y_{L} \le \hat{y} \le y^*+y_{U}.
$$
Then useful $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.
As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
In this case, joint inference will either estimate $\hat{w}$ to have fidelity 1 or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.

@@

Previous work in phylogenetics has developed consistency counterexamples using the same four-taxon topologies used here \cite{Felsenstein1978-rr}.
In this previous work, when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more optimal when performing a particular estimation procedure.
This is the converse of what we have shown for joint inference, where the Felsenstein topology is more likely than the Farris topology.
Moreover, the inconsistency in \cite{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
This is not the case here; for our case, the inconsistency generally occurs when all branches are long, and has more to do with the choices of the form of the likelihood than from the interplay of long and short branches.
Difficulties in phylogenetic estimation when generating data on the ``Farris zone'' tree have been found by \cite{Siddall1998-hq}, though \cite{Swofford2001-hr} show that sequence length plays a major role in these issues.
